---
title: "PlantGrowth - Data handling and visualisation"
output: html_notebook
date: '2022-10-12'
---

## Load PlantGrowth dataset

```{r}
data("PlantGrowth")
data("ToothGrowth")
```


## Explore the data frame

Use the functions you learned above to print out the column names of the PlantGrowth data set. Then print out the column with the weights.

```{r}
dim(PlantGrowth)
names(PlantGrowth)
head(PlantGrowth, 5)
```

## Manipulating data frames

* Filter the PlantGrowth data for experiments from the control group.
* Additionally select for weight.

```{r}
t <- PlantGrowth %>%
  filter(group=="ctrl") %>% 
  select("weight")
head(t, 3) # Showing only first three lines
```

## Plotting with ggplot2

Complete the code below to plot the plant weights against the groups.

```{r eval=FALSE}
PlantGrowth %>% 
  ggplot(   ) +
```

```{r}
PlantGrowth %>% 
  ggplot( mapping = aes(x = group, y = weight) ) +
  geom_point()
```

## Plotting categorical data

Plot the plant weights against the groups using a boxplot.

```{r}
PlantGrowth %>% 
  ggplot( mapping = aes(x = group, y = weight) ) +
  geom_boxplot()
```

## More than two variables

The `ToothGrowth` data set has three variables: A supplement, a dose, and the resulting tooth length.

```{r}
head(ToothGrowth, 3)
```

Complete the code below to plot the tooth length against dosage, while faceting by supplement.

```{r eval=FALSE}
ToothGrowth %>% 
  ggplot(aes(x=   , y=    ))+
  geom_point()+
  facet_wrap()
```

```{r eval=FALSE}
ToothGrowth %>% 
  ggplot(aes(x=dose, y=len))+
  geom_point()+
  facet_wrap(vars(supp))
```

## Continuous variables

Combine filtering and plotting: Filter for Adelie penguins first, then plot body mass against flipper length in a scatter plot.

```{r eval=FALSE}
library(palmerpenguins)
penguins %>% 
  filter(  ) %>% 
  ggplot(  ) +
```

```{r}
library(palmerpenguins)
head(penguins)
```

```{r}
penguins %>%
  drop_na() %>%
  filter( species == "Adelie" ) %>%
  ggplot( aes(x=body_mass_g, y=flipper_length_mm)) +
  geom_point()
```











