---
title: "Hypothesis testing"
output: html_notebook
date: '2022-10-12'
---

By the end of this third interactive tutorial you will be able to:

* explain the common principles behind statistical tests
* know how to spot common pitfalls
* perform simple tests in R

Before starting this tutorial, we recommend:

* you complete the first tutorial on (data handling and visualization)[https://www.ebi.ac.uk/training/online/courses/biostatistics-introduction/data-handling-and-visualisation/] or have basic knowledge or R and the tidyverse
* you complete the second tutorial on (probability distribution)[https://www.ebi.ac.uk/training/online/courses/biostatistics-introduction/probability-distributions/] or have basic knowledge of statistical distributions

## Overview and motivation

Prerequisites:
* Basic knowledge in R and tidyverse
* Basic knowledge on statistical distributions

Learning objectives:

* understand the common principles behind statistical tests
* learn to spot common pitfalls
* perform simple tests in R

Author: Sarah Kaspar, PhD

Contact: <sarah.kaspar@embl.de>

License: CC BY-NC-SA

Sources

* This tutorial is based on a lecture on testing by Bernd Klaus and Wolfgang Huber (2018).
* The demonstration in R roughly follows an example shown (here)[https://rafalab.github.io/pages/harvardx.html].
* The tutorial uses data from Winzell, M. S., & Ahrén, B. (2004).
* See the References section for more details.

### Steps of hypothesis testing

Let’s recapitulate what happens during hypothesis testing. Ideally, you undertake the following steps:

1. Set up a null model or null hypothesis
1. Collect data
1. Calculate the probability of the data in the null model
1. Decide: Reject the null model if the above probability is too small.

## Our first test: Disease prevalence

Here is what we have just seen:

* In a test group of persons with a precondition, 9 out of 100 have the disease of interest.
* We compare to a known prevalence of 4%, that means 4% of the entire population have the disease, when not distinguishing between persons with and without the precondition.
* We want to know whether the observed 9 out of 100 are significantly different from 4%. The null hypothesis thus is that they are not.
* The alternative hypothesis is that the prevalence in the test group differs from 4%.
* As we have 100 cases and a bimodal outcome (disease or no disease), a suitable null model is a binomial distribution with n=100 and success probability p=0.04.

### Binomial test in R

R is very convenient for calculating probabilities under different distributions, and also has functions for the common statistical tests.

As a reminder, functions in R that calculate numbers for the binomial distribution all end with binom, and the prefix specifies what to calculate:
* dbinom: density
* pbinom: cumulative distribution function (percentage of values smaller than)

For example, you can calculate the probability of observing 5 persons out of 100 with a disease, given a prevalence of 4%:

```{r}
n = 100 # number of test persons
p = 0.04 # given prevalence
dbinom(x=5, size=n, prob=p)
```

**Exercise:**

Can you perform the test that I showed you in the video, using pbinom?
For this, you need to calculate the probability of observing 9 or more persons with a disease out of 100, given a ''null prevalence'' of 4%.

```{r}
# we use q=8, because if lower.tail=FALSE, the probability P(X>x) is calculated.
# Check help(pbinom) for this.
pbinom(q=8, size=100, prob=0.04, lower.tail=FALSE)
```

Look up `binom.test`. This function performs a binomial test – which is what you have just done manually by calculating a binomial probability under a null hypothesis. Use this function to perform the test. Do you get the same result as with pbinom?

```{r}
binom.test(9,100,p=0.04)
```


















